\documentclass[sigconf]{acmart}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{listings}
\usepackage[noline]{algorithm2e}
\usepackage{comment}
\usepackage{color}
\usepackage{multirow}
\usepackage{balance}
\usepackage{framed}
\usepackage{alltt}
\usepackage{xcolor}


% TODO
% copyright
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}


% TODO
%\acmDOI{10.475/123_4}

% TODO
%\acmISBN{123-4567-24-567/08/06}

% TODO
%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
%\acmYear{2018}
%\copyrightyear{2018}

% TODO
%\acmArticle{4}
%\acmPrice{15.00} 



% configures style of source code listings
\lstset{
	basicstyle=\small\sffamily
	%basicstyle=\scriptsize\ttfamily
	%numbers=left
}

\lstdefinelanguage{java}{
	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
	sensitive=false,
	morestring=[b]",
	morestring=[b]',
	morecomment=[l]{//},
	morecomment=[s]{/*}{*/)},
	escapeinside=??,
	moredelim=[is][\textit]{__}{__},
	moredelim=[is][\textbf]{_*}{*_}
}
\lstnewenvironment{displayjava}
	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
\lstnewenvironment{java}
	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}


\newcommand{\code}[1]{{\sf \small #1}\xspace}

\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\mt}[1]{\mathtt{#1}}

\newcommand{\todo}[1]{{\bfseries TODO #1}}

% for writing comments
\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains

% coloring and font style commands
  % ANTLR
\newcommand{\antlrap}{\textquotesingle}
\newcommand{\antlrparserrule}[1]{\textcolor{blue}{#1}}
\newcommand{\antlrlexerrule}[1]{\textcolor{red}{#1}}
\newcommand{\antlrliteralnoap}[1]{\textcolor{brown}{#1}}
\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
\newcommand{\antlrregex}[1]{\textcolor{olive}{#1}}
  % MPS in general
\newcommand{\mpsconcept}[1]{\textcolor{olive}{#1}}
\newcommand{\mpsinterface}[1]{\textcolor{violet}{#1}}
  % MPS structure
\newcommand{\mpsstkeyword}[1]{\textcolor{blue}{\textbf{#1}}}
\newcommand{\mpsstalias}[1]{\textcolor{olive}{\textbf{#1}}}
\newcommand{\mpsstproperty}[1]{\textcolor{violet}{\textbf{#1}}}
\newcommand{\mpsstplaceholder}[1]{\textcolor{gray}{\textbf{#1}}}
\newcommand{\mpsstcardinality}[1]{\textcolor{cyan}{\textbf{#1}}}
  % MPS editor
\newcommand{\mpsedkeyword}[1]{\textcolor{black}{\textbf{#1}}}
\newcommand{\mpsedtarget}[1]{\textcolor{magenta}{\textbf{#1}}}
\newcommand{\mpsedannotation}[1]{\textcolor{gray}{\textbf{#1}}}
  % MPS textgen
\newcommand{\mpstgkeyword}[1]{\textcolor{blue}{\textbf{#1}}}
\newcommand{\mpstgtarget}[1]{\textcolor{magenta}{\textbf{#1}}}
\newcommand{\mpstgparam}[1]{\textcolor{violet}{\textbf{#1}}}
\newcommand{\mpstgaction}[1]{\textcolor{violet}{\textbf{#1}}}
\newcommand{\mpstgliteral}[1]{\textcolor{olive}{\textbf{#1}}}
\newcommand{\mpstgnodeprop}[1]{\textcolor{darkgray}{\textbf{#1}}}



\begin{document}


\title{INGRID: Creating Languages in MPS from ANTLR Grammars}


\author{P\v{r}emysl Vysok\'y}
\affiliation{
  \institution{Charles University, Faculty of Mathematics and Physics}
  \city{Prague}
  \country{Czech Republic}
}
\email{premek.vysoky@gmail.com}

\author{Pavel Par\'izek}
\affiliation{
  \institution{Charles University, Faculty of Mathematics and Physics}
  \city{Prague}
  \country{Czech Republic}
}
\email{parizek@d3s.mff.cuni.cz}

\author{V\'aclav Pech}
\affiliation{
  \institution{JetBrains}
}
\email{vaclav@jetbrains.com}



\begin{abstract}
JetBrains MPS is a language workbench, an IDE that allows developers both to write code and create language definitions.
It leverages the concept of projectional editing, where the developer directly manipulates the AST representation of program code.
While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPPLs) --- for example, to compile and run programs written in some DSL.
%, or to enable combination of GPPLs with DSLs created in MPS.

We present \textsc{Ingrid}, a method for construction of a language definition in MPS based on its ANTLR grammar.
The structure of a language is generated automatically, but its projectional editor and other aspects has to be adjusted manually.
%by the user.
During the development of \textsc{Ingrid}, we encountered several challenges related to the mapping between grammars and language definitions in MPS that are based on object-oriented principles.
Another difficulty is that grammars do not hold any information about code layout.

We implemented \textsc{Ingrid} as a plugin for MPS and evaluated it on several mainstream GPPLs, such as JavaScript and C\#.
Results show that our approach is practical, and it saves the user from many hours of tedious and error-prone work.
Necessary manual adjustments take only a short time.
\end{abstract}


\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011019</concept_id>
<concept_desc>Software and its engineering~Extensible languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041.10011047</concept_id>
<concept_desc>Software and its engineering~Source code generation</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041.10011688</concept_id>
<concept_desc>Software and its engineering~Parsers</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011066.10011069</concept_id>
<concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Extensible languages}
\ccsdesc[500]{Software and its engineering~Source code generation}
\ccsdesc[500]{Software and its engineering~Parsers}
\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}

\keywords{language workbench, projectional editor, grammar, domain-specific language, ANTLR, JetBrains MPS}

\maketitle



\input{chapters/introduction}
\input{chapters/background}
\input{chapters/language_definition}
\input{chapters/evaluation}
\input{chapters/futurework}


\begin{acks}
% TODO mozna pridat jeste nejaky dalsi grant (az clanek nekam vezmou)
This work was partially supported by JetBrains.
\end{acks}


\bibliographystyle{ACM-Reference-Format}

\begin{thebibliography}{}

\bibitem{ref:MPSBOOK} Fabien Campagne. 2014. The MPS Language Workbench, Volume I. CreateSpace Independent Publishing Platform.

\bibitem{ref:DHL80} Veronique Donzeau-Gouge, Gerard Huet, Bernard Lang, and Gilles Kahn. 1980. Programming Environments Based on Structured Editors: The Mentor Experience. Research Report RR-0026, INRIA.

\bibitem{ref:BISONFLEX} John Levine. 2009. Flex \& Bison: Text Processing Tools. O'Reilly Media.

\bibitem{ref:KV10} Lennart C.L. Kats and Eelco Visser. 2010. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. In Proc. of OOPSLA 2010, ACM.

\bibitem{ref:ANTLRBOOK} Terence Parr. 2013. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf.

\bibitem{ref:MBEDDR} Markus Voelter, Daniel Ratiu, Bernhard Schaetz, and Bernd Kolb. 2012. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. In Proc. of SPLASH 2012, ACM.

\bibitem{ref:VSB14} Markus Voelter, Janet Siegmund, Thorsten Berger, and Bernd Kolb. 2014. Towards User-Friendly Projectional Editors. In Proc. of SLE 2014, LNCS, 8706.

\bibitem{ref:VWK15} Markus Voelter, Jos Warmer, and Bernd Kolb. 2015. Projecting a Modular Future. IEEE Software, 32(5).

\bibitem{ref:TRFULL} Premysl Vysoky, Pavel Parizek, and Vaclav Pech. 2018. INGRID: Creating Languages in MPS from ANTLR Grammars (Full version). Tech. Report No. D3S-TR-2018-01, Dep. of Distributed and Dependable Systems, Charles University.

\bibitem{ref:CSHARP} C\# Language Specification, Version 5.0 (2012), Microsoft Corporation, \url{https://msdn.microsoft.com/en-us/library/ms228593.aspx}

\bibitem{ref:ECMASCRIPT51} ECMAScript Language Specification, Edition 5.1 (June 2011), \url{http://www.ecma-international.org/ecma-262/5.1/Ecma-262.pdf}

\bibitem{ref:JSON} The JSON Data Interchange Format, 1st Edition (October 2013), \url{http://www.ecma-international.org/publications/files/ECMA-ST/ECMA-404.pdf}

\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}

\bibitem{ref:ANTLR} ANTLR parser generator, \url{http://www.antlr.org}

\bibitem{ref:ANTLR2MPS} Fabien Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}

\bibitem{ref:MPSMETABNF} DSLFoundry. The mps-metabnf project, \url{https://github.com/DSLFoundry/mps-metabnf}

\bibitem{ref:PE4MPS} Marco Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}

\bibitem{ref:PE} Marco Lombardo. PE grammars, \url{https://github.com/mar9000/pe}

\bibitem{ref:BASELANG} MPS BaseLanguage, \url{https://confluence.jetbrains.com/display/MPSD34/Base+Language}

\end{thebibliography}

\input{chapters/appendix}

\end{document}

